---
title: "Announcing ggforce: Accelerating ggplot2"
description: "ggforce is a collection of extensions to the ggplot2 plotting system intended to add missing pieces, boost functionality, and break the boundaries of what's possible."
date: '2016-11-22'
tags: [R, package, announcement, ggforce, ggplot2]
categories: [R]
large_thumb: true
img:
    thumb: "/assets/images/ggforce_collage.png"
---

```{r, include=FALSE}
library(knitr)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
Sys.setenv(LANG = "en")

library(ggforce)
```

I am very pleased to announce the first release of the `ggforce` package. 
`ggforce` is a general `ggplot2` extension package in the same vein as [`ggalt`](https://CRAN.R-project.org/package=ggalt) with no overarching goal 
other than to provide additional functionality to the `ggplot2` universe. The 
inception and birth of ggforce was documented in a recent 
[post](`r blogdown::shortcode('ref', '\\"2016-10-31-Becoming-the-intern.md\\"')`) and the road to CRAN has 
been long and winding, with my personal goal of the package going through 
multiple iterations. At the beginning it was meant as a place where everyone
could submit extensions to. As I began working on 
[`ggraph`](https://github.com/thomasp85/ggraph) it became a dumping ground for 
functionality that was broader than network visualization, and as I began 
working on `ggplot2` it became a testing ground for new facetting functions. 
Now, as it is ready for CRAN, it does include some requests and code from other
developers, thus paying partial homage to the founding goal of the package.

This post will not go into detail with every functionality that ggforce adds to
`ggplot2`. `ggforce` includes a 
[vignette](https://cran.r-project.org/web/packages/ggforce/vignettes/Visual_Guide.html) 
where every feature is described along with motivation and code examples.

## Features
Below follows a description of all the major features (small utility functions
are not included), grouped by the grammar they extend:

### Geoms
* `geom_arc()` allows you to plot circle segments defined by center, radius, and
start- and end-angle.
* `geom_arc_bar()` is as above, except it draws a *thick* arc defined by an 
inner and outer radius. If the inner radius is zero it becomes a wedge.

```{r}
pie <- data.frame(
    state = c('eaten', 'eaten but said you didn\'t', 'cat took it', 
              'for tonight', 'will decompose slowly'),
    focus = c(0.2, 0, 0, 0, 0),
    start = c(0, 1, 2, 3, 4),
    end = c(1, 2, 3, 4, 2*pi),
    amount = c(4,3, 1, 1.5, 6),
    stringsAsFactors = FALSE
)
ggplot(pie) + 
    geom_arc_bar(aes(x0 = 0, y0 = 0, r0 = 0, r = 1, amount = amount, 
                     fill = state, explode = focus), stat = 'pie') + 
    scale_fill_brewer('', palette = 'Set1') +
    coord_fixed()
```

* `geom_circle()` makes it possible to draw circles based on coordinate space,
that is, the radius is scaled to the coordinate system rather than absolute.
* `geom_link()` is like `geom_segment()` except it expands the line into 
multiple segments making it possible to draw gradients over the line.
* `geom_bezier()` allows you to draw cubic and quadratic beziers.

```{r}
beziers <- data.frame(
    x = c(1, 2, 3, 4, 4, 6, 6),
    y = c(0, 2, 0, 0, 2, 2, 0),
    type = rep(c('cubic', 'quadratic'), c(3, 4)),
    point = c('end', 'control', 'end', 'end', 'control', 'control', 'end')
)
help_lines <- data.frame(
    x = c(1, 3, 4, 6),
    xend = c(2, 2, 4, 6),
    y = 0,
    yend = 2
)
ggplot() + geom_segment(aes(x = x, xend = xend, y = y, yend = yend), 
                        data = help_lines, 
                        arrow = arrow(length = unit(c(0, 0, 0.5, 0.5), 'cm')), 
                        colour = 'grey') + 
    geom_bezier(aes(x= x, y = y, group = type, linetype = type), 
                data = beziers) + 
    geom_point(aes(x = x, y = y, colour = point), data = beziers)
```

* `geom_bspline()` makes it possible to draw b-splines.
* `geom_sina()` is a novel alternative to `geom_violin()`/`geom_boxplot()`/`geom_jitter()`
that was submitted to ggforce by the developers of 
[sinaplot](https://CRAN.R-project.org/package=sinaplot)

```{r}
df <- data.frame(
  "Distribution" = factor(c(rep("Unimodal", 500),
                     rep("Bimodal", 250),
                     rep("Trimodal", 600)), c('Unimodal', 'Bimodal', 'Trimodal')),
  "Value" = c(rnorm(500, 6, 1),
              rnorm(200, 3, .7), rnorm(50, 7, 0.4),
              rnorm(200, 2, 0.7), rnorm(300, 5.5, 0.4), rnorm(100, 8, 0.4))
)
ggplot(df, aes(Distribution, Value)) + 
    geom_sina(aes(colour = Distribution), size = 1)
```

### Scales
* `scale_[x|y]_unit()` provides support for the units class from the [units](https://CRAN.R-project.org/package=units) package. It provides 
automatical labelling of axis with units as well as changing the units of the 
scale without touching the data.

```{r}
library(units)
miles <- make_unit('miles')
gallon <- make_unit('gallon')
horsepower <- make_unit('horsepower')
mtcars$consumption <- mtcars$mpg * (miles/gallon)
mtcars$power <- mtcars$hp * horsepower

ggplot(mtcars) +
    geom_point(aes(power, consumption)) +
    scale_y_unit(unit = 'km/l')
```

### Facets
* `facet_[grid|wrap]_paginate()` allows you to split grid and wrap facetting out
into multiple pages.
* `facet_zoom()` is a third zooming alternative that provides a context 
overview.

```{r}
ggplot(iris, aes(Petal.Length, Petal.Width, colour = Species)) +
    geom_point() +
    facet_zoom(x = Species == "versicolor")
```

### Transformations
While transformations are part of the `scales` package rather than `ggplot2` 
their use is integral to `ggplot2`, and `ggforce` thus fills some gaps in this
department as well.

* `power_trans()` lets you create any power transformation (`scales` only 
provides `sqrt_trans()`).
* `radial_trans()` lets you translate between coordinates specified by radius 
and angle and coordinates specified by x and y.
* `trans_reverser()` reverses any monotonous transformation function, making it
possible to e.g. have a reversed log transformation (`scales` only provides
`reverse_trans()` for reversing the identity transformation).

```{r}
p3 <- power_trans(3)
p3r <- trans_reverser(p3)
ggplot(mtcars) + 
    geom_point(aes(mpg, cyl)) + 
    scale_y_continuous(trans = p3r)
```

## Future
I have many cool ideas planned for future releases, and I hope that users will
approach me if they have some great ideas as well. If you have a burning wish 
for something in the `ggplot2` ecosystem I welcome any issues and PRs on the
[`ggforce` github repository](https://github.com/thomasp85/ggforce). The 
`ggplot2` ecosystem is thriving though so please visit 
<http://www.ggplot2-exts.org> and see if your idea doesn't already exist in 
another package.

Happy plotting!
