---
title: "Data Driven Logo Design with ggraph and gtable"
description: "I recently created the logo for www.data-imaginist.com and in this post I will describe how I did it and what went through my head in the process"
date: '2016-10-07'
slug: 'Data-driven-logo'
tags: [R, ggraph, design]
categories: [R]
large_thumb: true
img:
    thumb: "/assets/images/logo_large.png"
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE
)
Sys.setenv(LANG = "en")

library(igraph)
library(ggraph)
library(gtable)
library(grid)
d_graph <- readRDS('../../../../Graphics/di_logo_data.rds')$d_graph
```

After having "succesfully" [launched my blog](`r blogdown::shortcode('ref', '\\"2016-09-18-wellcome-to-data-imaginist.md\\"')`)
I felt the time had come to create some identity for it (and by extension, me). 
What better way to do this than create a logo? This thought didn't suddenly 
appear in my mind - if you've read the announcement post you will now that such
details have been rummaging around my head for as long as the thought of a blog.

My initial idea was to make something very stylistic in the vein of a *d i* 
ligature in a classic serif font - I have absolutely no idea where the thought
came from, but I think it was inspired by writing my PhD dissertation with the 
[EB Garamond](http://www.georgduffner.at/ebgaramond/) font. Something about the
idea didn't click though: *Data Imaginist* and then a typography-based logo? No,
it had to convey a sense of playfulness with data and visualization! I still 
wanted a *D* and an *I* somehow, so the question ended up with how to create 
data visualizations in the shape of letters.

## Drawing a *D*
My instinct with the D was to make a sort of horizontal bar chart that roughly
outlined the shape of a D. I kind of felt that this was a bit too boring and 
unimaginative though, so I continued to look. Something got my thinking of an 
arc diagram and how the correct graph structure could actually produce a pretty 
decent D, with a hole and everything. Such a graph structure would be hard to 
find in the wild though so I had to make one myself.

```{r, eval=FALSE}
# Create a random graph with 100 nodes and 1000 edges
d_data <- data.frame(from = sample(100, 1000, T), to = sample(100, 1000, T))
# Remove all loops
d_data <- d_data[d_data$from != d_data$to, ]
# Find all edges that connects the top and bottom (the large arc)
d_data_top <- d_data[abs(d_data$from - d_data$to) >= 75, ]
# Find a subset of edges that only connects nodes close to each other (the stem)
d_data_bottom <- d_data[sample(which(abs(d_data$from - d_data$to) <= 25), 200), ]
# Assemble it all to an igraph object
d_graph <- graph_from_edgelist(as.matrix(unique(rbind(d_data_top, d_data_bottom))))
# Add a random class to each node
V(d_graph)$class <- sample(letters[1:10], length(unique(unlist(d_data))), T)
```

Bear in mind that a lot of parameter tweaking went into finding a nice looking
graph. Also, keep in mind that `sample` is used in a few places so if you run
this you will end up with a slightly different graph. I never recorded the seed
I used when I created the logo so my version will forever be unique.

With the data ready it was time to draw the D. Thankfully I had already
implemented an arc diagram layout in [ggraph](https://github.com/thomasp85/ggraph) 
so it should be a piece of cake. In order to add some flair to the plot I 
decided to draw the arcs with an opacity gradient from start to end:

```{r}
D <- ggraph(d_graph, layout = 'linear') +
    geom_edge_arc(aes(x=y, y=x, xend=yend, yend=xend, alpha = ..index..,
                      color = node2.class))
D
```

Hmm, there is the idea of a *D* somewhere in that plot but it was not what I had
envisioned. The "problem" is that edges going in different directions are 
positioned on different sides of the y-axis. I could of course change the graph 
so all edges went in the same direction, but then the gradient would also go in 
the same direction for each arc, which was not what I wanted. The nice thing 
about being the developer behind your own tools is that you can always make them
do your biding, so I promptly added a `fold` argument to `geom_edge_arc()` that
would put all arcs on the same side irrespectively of their direction (If you
are ever to use this feature, be thankful that I had to draw a D).

```{r}
D <- ggraph(d_graph, layout = 'linear') +
    geom_edge_arc(aes(x=y, y=x, xend=yend, yend=xend, alpha = ..index..,
                      color = node2.class), 
                  fold = TRUE)
D
```

Now we're getting somewhere, but lets loose the legends and coordinate system 
and make the arcs a bit thicker while we're at it...

```{r}
theme_empty <- theme_void() +
    theme(legend.position = 'none', 
          plot.margin = margin(0, 0, 0, 0, 'cm'), 
          legend.box.spacing = unit(0, 'cm'))

D <- ggraph(d_graph, layout = 'linear') +
    geom_edge_arc(aes(x=y, y=x, xend=yend, yend=xend, alpha = ..index..,
                      color = node2.class),
                  fold = T, edge_width = 1) +
    scale_y_continuous(expand = c(0.03,0)) +
    theme_empty
D
```

# Drawing an *I*
The shape of an *I* is quite easier to make with a visualization, but it is in
turn also more boring so the visualization would need to be a tad more 
interesting to make up for it. I decided to go for a lowercase *i* as it would
allow my to do different things with both the dot and the stem. Already being in
ggraph-land after the D I decided to continue down that road. A treemap would be
a nice fit for the rectangular stem, while any circular layout would work well 
for the dot. The only thing to keep in mind for the treemap was that since it 
was meant to drawn as a very thin rectangle, the aspect ratio should be set in 
the layout to make sure the rectangles remains fairly square.

We don't need to make up any data for the *i* as ggraph already comes with a 
hierarchical data structure we can use, namely the [flare](http://flare.prefuse.org) 
class hierarchy, so lets get right to it:

```{r}
# Create an igraph object from the flare data
flareGraph <- graph_from_data_frame(flare$edges, vertices = flare$vertices)
# Set the class of each node to the name of their topmost-1 parent
flareGraph <- tree_apply(flareGraph, function(node, parent, depth, tree) {
    tree <- set_vertex_attr(tree, 'depth', node, depth)
    if (depth == 1) {
        tree <- set_vertex_attr(tree, 'class', node, V(tree)$shortName[node])
    } else if (depth > 1) {
        tree <- set_vertex_attr(tree, 'class', node, V(tree)$class[parent])
    }
    tree
})
# Define wether a node is terminal
V(flareGraph)$leaf <- degree(flareGraph, mode = 'out') == 0
```

With our graph at hand we can draw the stem:

```{r}
i_stem <- ggraph(flareGraph, 'treemap', weight = 'size', width = 1, height = 3) +
    geom_node_tile(aes(filter = leaf, fill = class, alpha = depth), colour = NA) +
    geom_node_tile(aes(filter = depth != 0, size = depth), fill = NA) +
    scale_alpha(range = c(1, 0.3), guide = 'none') +
    scale_size(range = c(1.5, 0.4), guide = 'none') +
    theme_empty
i_stem
```

Well, it looks decidedly non-thin, but this is just a matter of stretching it in
the right direction...

For the dot we can continue with our flare graph and draw a circular hierarchy,
but since we have same additional information about the flare data (which 
classes imports each other) we can do something more fancyfull by drawing these
imports as [hierarchical edge bundles](http://ieeexplore.ieee.org/document/4015425/).
Hierarchical edge bundles is a way to bundle connections by letting them
loosely follow an underlying hierarchy in the data structure and it makes for 
some very pretty plots.

```{r}
importFrom <- match(flare$imports$from, flare$vertices$name)
importTo <- match(flare$imports$to, flare$vertices$name)
i_dot <- ggraph(flareGraph, 'dendrogram', circular = TRUE) +
    geom_conn_bundle(aes(colour = ..index..), data = get_con(importFrom, importTo),
                     edge_alpha = 0.25) +
    geom_node_point(aes(filter = leaf, colour = class)) +
    coord_fixed() +
    theme_empty
i_dot
```

What a nice dot!

The last touch before we assemble it all is to apply a less ggplot2-y colour
scale. I have absolutely no perfect method to this. I like to scavenge places 
like [Adobe Color CC](https://color.adobe.com/da/explore/most-popular/?time=all)
but these palettes are mainly for design work and often only provides <=5 
different colours (which is often too few for data visualizations). This time I
decided to expand on one of the palettes 
([Flat design colors 1](https://color.adobe.com/da/Flat-design-colors-1-color-theme-3044245/))
by adding darker copies of each color to it. In the end the palette ended up
like this

```{r, fig.show='hold', out.width='250px', out.height='250px'}
palette <- paste0('#', c('2B6E61', 'AB9036', '99532B', '9C3F33', '334D5C', 
                         '45B29D', 'EFC94C', 'E27A3F', 'DF5A49', '677E52'))

D <- D + scale_edge_color_manual(values = palette)
i_stem <- i_stem + scale_fill_manual(values = palette)
i_dot <- i_dot + scale_edge_colour_gradient('', low = 'white', high = '#2C3E50') +
    scale_color_manual(values = palette)
print(D)
print(i_stem)
print(i_dot)
```

# The Assembly
[gtable](https://github.com/hadley/gtable) is an underappreciated part of the 
whole ggplot2 experience as user rarely know that it is this layout engine that 
is used to position all those nice geoms. While it is rarely used outside of
ggplot2 it does not have to be so.

At its simplest gtable is a way to define a grid with varying widths and heights
of each column/row and add content that spans one or multiple rows/columns. For
our logo we want a wide column for the *D* and a thinner column for the *i* as
well as some spacing. The dot needs to be placed in a square cell so we need to 
have a row with the same height as the width of the *i* column. In the end we 
(through some trial and error) ends up with this grid:

```{r, echo=FALSE, fig.asp=1}
gtable_show_layout(
    gtable(unit(c(1.4, 0.15, 0.6, 0.15), 'null'), 
           unit(c(0.15, 0.6, 0.15, 1.4), 'null'), respect = TRUE))
```

Before we can put our letters into the grid, they need to be converted into
something that gtable (or grid) can understand. Thankfully ggplot2 exports this
functionality in the form of the `ggplotGrob()` function.

```{r}
D_table <- ggplotGrob(D)
i_dot_table <- ggplotGrob(i_dot)
i_stem_table <- ggplotGrob(i_stem)
composite <- gtable(widths = unit(c(1.4, 0.15, 0.6, 0.15), 'null'), 
                    heights = unit(c(0.15, 0.6, 0.15, 1.4), 'null'), 
                    respect = TRUE)
composite <- gtable_add_grob(composite, D_table, 1, 1, 4, 2)
composite <- gtable_add_grob(composite, i_dot_table, 1, 2, 3, 4)
composite <- gtable_add_grob(composite, i_stem_table, 4, 3)
grid.newpage()
grid.draw(composite)
```

É viola! A (sort of) reproducible logo made entirely in R. In a future post I 
will investigate how we can add some animation to it...