---
title: "Announcing ggraph: A grammar of graphics for relational data"
description: "ggraph, a package for creating network and tree visualizations using the ggplot2 API has just been released on CRAN"
date: '2017-02-23'
slug: 'Announcing-ggraph'
tags: [R, package, announcement, ggraph, ggplot2, visualization]
categories: [R]
large_thumb: true
img:
    thumb: "/assets/images/ggraph_announce.png"
---

```{r, include=FALSE}
library(knitr)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
Sys.setenv(LANG = "en")

library(ggforce)
```

I am absolutely thrilled to announce that `ggraph` has finally been released
on CRAN. `ggraph` is my most ambitious package to date and its very early 
genesis has been described in a [prior post](`r blogdown::shortcode('ref', '\\"2016-10-31-Becoming-the-intern.md\\"')`).
If any mention of `ggraph` is completely new to you, then in short terms `ggraph`
is an extension of the `ggplot2` API to support relational data such as networks
and trees. I feel fairly confident in saying that `ggraph` is the most
powerful way to create static network based visualizations in R. Leading up to 
the release, the three main concepts of `ggraph` has been described in detail in 
their own blog posts 
([layouts](`r blogdown::shortcode('ref', '\\"2017-02-06-ggraph-introduction-layouts.html\\"')`), 
[nodes](`r blogdown::shortcode('ref', '\\"2017-02-10-ggraph-introduction-nodes.html\\"')`), and 
[edges](`r blogdown::shortcode('ref', '\\"2017-02-16-ggraph-introduction-edges.html\\"')`)) so this
will not be reiterated here. Instead I'll talk a bit about the philosophy behind
the package as well as show of some of the features that do not fall into any of
the three main concepts.

## The Philosophy
There is no shortage of software for creating network visualizations and there
is no shortage of said visualizations themselves. Often though, the 
visualizations are more impressive than informative and it is easy to feel that
their main task is to show that we are really dealing with some complex data.
All of this has led to a certain disdain for classic network visualizations
perfectly encapsulated in the nickname *hairballs*. It does not have to be like 
this! The greatness of `ggplot2` lies in how it allows users to quickly iterate
over visualization approaches, thus better ensuring that the best visualization
approach is reached. If this was extended to relational data it is my belief
that users would be more likely to try to make plots that are more meaningful. 
After all we all want interpretability, right? Consider having to try out 7 
different network visualization packages with different APIs versus just mixing
and matching layouts and geoms in an iterative process --- I know which way I 
prefer.

The goal of `ggraph` is thus clear --- provide everything related to 
visualizations of relational data in a `ggplot2`-like API to lessen the 
cognitive load on experimenting with different visual representations. I'm not
there yet, but I feel the current version represents a solid foundation where
most users will not feel many limitations --- on the contrary I believe most
users will feel like the chains have come off and they are set free.

### Future focus
As I pointed out, `ggraph` is far from done. I'll try to keep my development
focus in the open by putting things on the road-map as 
[GitHub issues](https://github.com/thomasp85/ggraph/issues). Honorable mentions
include matrix, d3-force and sankey layout, expanded support for edge endings 
(more choices than `grid::arrow()` provides), edge routing (avoid node 
collision), and textbox nodes. I welcome all suggestions as the world of network
visualizations is moving fast and I cannot keep on top of everything.

## Features besides layouts, nodes, and edges
Understanding the node and edge geoms along with how layouts are defined will
get you a long way towards visualizing networks. Still, `ggraph` has more to 
offer, some of which will be discussed here:

### theme_graph()
Consider the following plot:

```{r, message=FALSE}
library(ggraph)
library(igraph)
graph <- graph_from_data_frame(highschool)

p <- ggraph(graph, layout = 'kk') + 
    geom_edge_link(aes(colour = factor(year))) + 
    geom_node_point() + 
    ggtitle('An example')
    
p
```

While the `ggplot2` heritage clearly shows due to the grey background with white
grid lines, the whole concept of x and y axes is often redundant in network
visualizations and are just a distraction. `ggraph` provides its own theme
optimized for network visualizations called `theme_graph()`, that facilitates 
clean and beautiful visualizations:

```{r}
p + theme_graph()
```

`theme_graph()`, besides removing axes, grids, and border, changes the font to
*Arial Narrow* (this can be overridden). Furthermore, it makes it easy to change
the coloring of the plot:

```{r}
p + theme_graph(background = 'grey20', text_colour = 'white')
```

Adding the same theme to every plot is tedious and `ggraph` provides a way to
avoid this. Using `set_graph_style()` the `theme_graph()` is set as default. As
an extra benefit all text-based geoms gets their defaults updated so the text
automatically uses the same style as the theme.

```{r}
set_graph_style()

p
```

### Facetting
A powerful but underutilized way of gaining insight into networks is by using 
small multiples. This technique can reduce edge over-plotting in a very 
meaningful way by spreading nodes and edges out based on their attributes. The
benefits of small multiples are not unique to relational data, as the popularity
of `ggplot2`s facetting functionality shows. The base facetting functions 
provided by `ggplot2` is a bad fit for networks though, as we are working with
two very distinct types of data. If you facet on a node attribute, all edges 
would be plotted in all panels, despite the terminal nodes not being present 
which is not what you expect. Because of this `ggraph` comes with its own set
of facetting functions tailored to network data:

#### facet_nodes() and facet_edges()
These two functions are equivalent to `facet_wrap()` in functionality, but they
only address node and edge data respectively. When using `facet_nodes()` edges
are only drawn in a panel if both terminal nodes are present there. When using
`facet_edges()` nodes are always drawn in all panels even if the node data
contains an attribute named the same as the one used for the edge facetting.

```{r}
# Assign each node to a random class
V(graph)$class <- sample(letters[1:4], gorder(graph), TRUE)
# Make year a character
E(graph)$year <- as.character(E(graph)$year)

p <- ggraph(graph, layout = 'kk') + 
    geom_edge_fan(aes(alpha = ..index.., colour = year)) + 
    geom_node_point(aes(shape = class)) + 
    scale_edge_alpha(guide = 'none')

p + facet_edges(~year)
```

Often, when working with small multiples it is nice to have some visual 
separation between each plot --- setting a foreground color in `theme_graph()`
will add strip background and border (you can also use the `th_foreground()` 
helper for this):

```{r}
p + facet_nodes(~class) + th_foreground(foreground = 'grey80', border = TRUE)

# Lets not have to add this everytime
set_graph_style(foreground = 'grey80')
```

#### facet_graph
Facetting on two variables simultaneously is very powerful and something that
is supported in `ggplot2` with `facet_grid()`. In `ggraph` the same is possible
using `facet_graph()` that takes the behavior of `facet_nodes()` and 
`facet_edges()` and combines them:

```{r}
p + facet_graph(year ~ class)
```

As with `facet_grid()` marginal plots are supported as well:

```{r}
p + facet_graph(year ~ class, margins = TRUE)
```

While the default is to put facet the rows on edges and the columns on nodes, 
this is free to change using the `row_type` and `col_type` arguments. There is
nothing stopping you from facetting on the same type in each dimension either:

```{r}
# Facet edge by the class of their start node as well as year
p + facet_graph(year ~ node1.class, col_type = 'edge')
```

I hope I have convinced you that facetting in the context of relational data is
both very easy, as well as extremely powerful. Avoiding the hairball is one of
the prime goal of network visualizations and using small multiples is a 
fantastic way of cutting down on the number of nodes and edges while still 
getting the full picture.

